---
output:
  html_document: default
  pdf_document: default
---
Discerning Prosper Interest Rates by Abhishek Anirudhan
========================================================
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```


```{r, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(tidyr)
library(data.table)
library(gridExtra)
library(GGally)
library(lubridate)
library(memisc)
library(RColorBrewer)
```

```{r, Load_the_Data}
# Load the Data

#Also replaced empty cells with NA

pld <- read.csv("D:/Udacity - Data Analyst/p4/prosperLoanData.csv", 
                na.strings = c('', ' ', 'NA'))

# And setting the theme

theme_set(theme_minimal())
```

### INTRODUCTION:

I will be examining a dataset of loans generated by the peer-to-peer lending forum, Prosper between 2006 and 2014. The dataset contains over 100,000 observations described over 81 features.I took up this project with the intention of probing the dataset to discern any interesting patterns and explain them.


#### LOOKING AT MISSING VALUES:

The dataset seemed to contain many NAs, I made some preliminary plots to find out just how many.     
         

```{r echo=FALSE, Univariate_Plots, fig.height=9, fig.width=9}

#Taking a look at NAs in the dataset

missing.Vals <- apply(pld, 2, is.na) 
missing.Vals <- apply(missing.Vals, 2, sum)
missing.Vals <- setDT(data.frame(missing.Vals), keep.rownames = TRUE)
names(missing.Vals) <- c('name','count')

#Subsetting all columns with more than 600 NAs

missing.Vals <- subset(missing.Vals, count != 0 & count > 600)

ggplot(data = missing.Vals, aes(x = name, y = count)) + 
  geom_bar(stat = 'identity', fill = 'darkblue') +
  xlab("Number of Observations") + coord_flip()
```
         
```{r echo=FALSE}
ggplot(data = subset(missing.Vals, count > 10000), aes(x = name, 
                                                       y = count)) + 
  geom_bar(stat = 'identity', fill = "darkblue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("More than 10000 NAs") +
  xlab("Variable Name") + ylab("Number of Observations")
  
```
           
I then proceed to remove these and other variables.     
            
```{r echo=FALSE}
#Removing colums with too a high number of missing values and other columns to be left out of the analysis

pld2 <- pld[, !(names(pld) %in% missing.Vals$name)]

pld2[,c("LP_GrossPrincipalLoss", "LP_CollectionFees", "LP_NetPrincipalLoss", 
        "LP_NonPrincipalRecoverypayments", "PercentFunded", "Recommendations", 
        "InvestmentFromFriendsCount", "InvestmentFromFriendsAmount", 
        "Investors", "LoanCurrentDaysDelinquent", "MemberKey", "LoanKey", 
        "ListingKey", "DateCreditPulled", "OpenRevolvingAccounts", 
         "OpenRevolvingMonthlyPayment", "LoanNumber")] <- NULL

# Adding two variables

pld2$EmploymentStatus <- pld$EmploymentStatus

# Restructuring the variable into a factor with labels

pld2$LoanPurpose <- factor(pld$ListingCategory..numeric., levels = c(0:20), 
                           labels = c("Not available", "Debt consolidation", 
                                      "Home improvement", "Business", 
                                      "Personal loan", "Student use", "Auto", 
                                      "Other", "Baby and adoption", "Boat", 
                                      "Cosmetic procedure", "Engagement ring", 
                                      "Green loans", "Household expenses", 
                                      "Large purchases", "Medical/dental", 
                                      "Motorcycle", "RV", "Taxes", 
                                      "Vacation", "Wedding loans"))
```
           
               
--------         

# Univariate Plots Section        

            
The original dataset contained 113,937 observations spread over 81 variables.         
                  
```{r}
dim(pld)
```
           
                 
This was reduced to 27 variables for the purposes of analysis. I thought a smaller datasets with not many missing points will be conducive to analysis. I might borrow excluded variables from the original dataset in the future, but for now I will start here:          
                  
```{r}
dim(pld2)
```

```{r}
names(pld2)
```
                         
                      
----------     

### BORROWER'S PROFILE                         

I will begin by exploring the profile of borrowers. As stated the dataset is composed on 113,937 borrowers.           
                                 
                                 
#### INCOME RANGE:                 
                                   
```{r echo=FALSE}
table(pld2$IncomeRange)
```
```{r echo=FALSE}
ggplot(pld2, aes(x = IncomeRange)) + geom_bar(fill = "lightpink") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Distribution of Borrowers by Income") +
  ylab("Number of Observations") +
  xlab("Income Range")
```

Right away, I can see that the data is not order, I correct for this, and plot the distribution again:

```{r echo=FALSE}

#Ordering the varibale

pld2$IncomeRange <- ordered(pld2$IncomeRange, 
                            levels = c("Not employed", "$0", "$1-24,999", 
                                       "$25,000-49,999","$50,000-74,999", 
                                       "$75,000-99,999", "$100,000+", 
                                       "Not displayed"))

head(pld2$IncomeRange)
```

```{r echo=FALSE}
ggplot(pld2, aes(x = IncomeRange)) + geom_bar(fill = "lightpink") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Number of Observations") +
  xlab("Income Range") +
  ggtitle("Distribution of Borrowers by Income (Ordered)")
```


Plotting only declared incomes (USD0 - USD100,000+), I can see that the distribution is approximately normal with most borrowers having incomes close to the middle (between USD25,000 and USD75,000)  


```{r echo=FALSE}
ggplot(subset(pld2, IncomeRange %in% c("$0", "$1-24,999", 
                                       "$25,000-49,999","$50,000-74,999", 
                                       "$75,000-99,999", "$100,000+")), 
       aes(x = IncomeRange)) + geom_bar(fill = "lightpink") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Observations") +
  xlab("Income Range") +
  ggtitle("Distribution of Borrowers by Income")
```

  
#### MONTHLY INCOMES: 

```{r echo=FALSE}
summary(pld2$StatedMonthlyIncome)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(pld2, aes(x = StatedMonthlyIncome)) + 
  geom_histogram(fill = "lightsteelblue") +
  ylab("Number of Observations") +
  xlab("Monthly Income") +
  ggtitle("Distribution of Incomes by Monthly Income")
```
                    
I can see here that the presence of outliers is throwing the distribution off. To correct for this, I will take a look at the distribution without the outliers.        
                      
```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(pld2, aes(x = StatedMonthlyIncome)) + 
  geom_histogram(fill = "paleturquoise") + 
  xlim(0, quantile(pld2$StatedMonthlyIncome, 0.9)) + 
  ggtitle("0.9 quantile") + 
  ylab("Number of Loans") +
  xlab("Income")

p2 <- ggplot(pld2, aes(x = StatedMonthlyIncome)) + 
  geom_histogram(fill = "paleturquoise") + 
  xlim(0, quantile(pld2$StatedMonthlyIncome, 0.99)) + 
  ggtitle("0.99 quantile") + 
  ylab("Number of Loans") +
  xlab("Income")

p3 <- ggplot(pld2, aes(x = StatedMonthlyIncome)) + 
  geom_histogram(fill = "paleturquoise") + 
  xlim(0, quantile(pld2$StatedMonthlyIncome, 0.999)) + 
  ggtitle("0.9990 quantile") + 
  ylab("Number of Loans") +
  xlab("Income")

p4 <- ggplot(pld2, aes(x = StatedMonthlyIncome)) + 
  geom_histogram(fill = "paleturquoise") + 
  xlim(0, quantile(pld2$StatedMonthlyIncome, 0.9995)) + 
  ggtitle("0.9995 quantile") + 
  ylab("Number of Loans") +
  xlab("Income")

grid.arrange(p1, p2, p3, p4)

```
                           
I observe that data upto the 90th quantile is approximately normal, but overall the data has a positive skew. This is expected because most borrowers - especially since they're borrowers, and hence need tp take out a loan for whatever purpose - have lower incomes.              

#### EMPLOYMENT STATUS:          

I see most of the people who are granted loans are employed.            
                           
```{r echo=FALSE}
table(pld2$EmploymentStatus)
```
```{r echo=FALSE}
ggplot(pld2, aes(x = EmploymentStatus)) + geom_bar(fill = "skyblue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Number of Loans") +
  xlab("Employment Status") +
  ggtitle("Borrowers by Employment Status")
```
                    
#### HOUSE OWNERSHIP:    
           
About equal number of loans were given to Homeowners as opposed to Non-Homeowners.      
                             
```{r echo=FALSE}
table(pld2$IsBorrowerHomeowner)
```
```{r echo=FALSE}
ggplot(pld2, aes(x = IsBorrowerHomeowner)) + geom_bar(fill = "cornflowerblue") + 
  ylab("Number of Loans") + 
  xlab("House Ownership") +
  ggtitle("Borrowers by Home Ownership")
```

                    
#### CREDIT SCORE:  

I see that I have a range (Lower and Upper) of credit scores. Converting these ranges to a single value will be fruitful for further analysis. I have done the same here:
```{r echo=FALSE}
#Calculated mean credit score from the lower and upper bounds of Credit Scores:

pld2 <- pld2 %>% mutate(CreditScoreMean = (CreditScoreRangeLower + 
                                             CreditScoreRangeUpper) / 2)

head(pld2$CreditScoreMean)
```

                     
```{r echo=FALSE}
summary(pld2$CreditScoreMean)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(pld2, aes(x = CreditScoreMean)) + geom_histogram(fill = "lavender") + 
  scale_x_continuous(breaks = seq(0, 900, 100)) + 
  ylab("Number of Loans") +
  xlab("Credit Score") +
  ggtitle("Loans by Credit Score")
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(pld2, aes(x = CreditScoreMean)) + 
  geom_histogram(fill = "lavender", breaks = seq(400, 900, 25)) + 
  xlim(400, 900) + scale_x_continuous(breaks = seq(0, 900, 50)) + 
  ylab("Number of Loans") +
  xlab("Credit Score") +
  ggtitle("Loans by Credit Score")
```

The distribution of credit scores is approximately normal. 
           
----------

### LOAN PORTFOLIO:

I will begin by plotting the distribution of the loan amounts. I observe that the distribution has a positive skew:    
        
```{r echo= FALSE}
summary(pld2$LoanOriginalAmount)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = pld2, aes(x = LoanOriginalAmount)) + 
  geom_histogram(fill = "thistle") + 
  geom_freqpoly() + 
  scale_x_continuous(breaks = seq(0, 40000, 5000)) + 
  ylab("Number of Loans") +
  xlab("Loan Amount") +
  ggtitle("Distribution of Loan Amounts")
```
```{r echo=FALSE, warning= FALSE, message= FALSE}
ggplot(data = pld2, aes(x = LoanOriginalAmount)) + 
  geom_histogram(fill = "thistle") + scale_y_log10() + 
  scale_x_continuous(breaks = seq(0, 40000, 5000)) + 
  ylab("Number of Loans") +
  xlab("Loan Amount") +
  ggtitle("Distribution of Loan Amounts")
```

#### LOAN STATUS:    

```{r echo=FALSE}
table(pld2$LoanStatus)
```

```{r echo=FALSE}
ggplot(data = pld2, aes(x = LoanStatus)) + geom_bar(fill = "lightsteelblue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Number of Loans") +
  xlab("Loan Status") +
  ggtitle("Distribution of Loan Status")
```

```{r echo=FALSE}
ggplot(data = pld2, aes(x = LoanStatus)) + geom_bar(fill = "lightsteelblue") + 
  scale_y_log10() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Number of Loans") +
  xlab("Loan Status") +
  ggtitle("Distribution of Loan Status")
```

#### LOAN TERM:

I see that most people either apply for or are granted loans for a duration of 36 months or 3 years, followed by durations of 60 months or 5 years, and 12 months.     
```{r echo=FALSE}
table(pld2$Term)
```
```{r echo=FALSE}
ggplot(data = pld2, aes(x = factor(Term))) + geom_bar(fill = "palevioletred") + 
  ylab("Number of Loans") +
  xlab("Loan Term") +
  ggtitle("Distribution of Loan Term")
```

#### LOAN PURPOSE:      

I see that most people borrow for Debt Consolidation. (I was pretty surprised at some of purposes listed!)
```{r echo = FALSE}
summary(pld2$LoanPurpose)
```
```{r echo = FALSE}
ggplot(data = subset(pld2, LoanPurpose != "Not available"), 
       aes(x = LoanPurpose)) + 
  geom_bar(fill = "lightgreen") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Number of Loans") +
  xlab("Loan Purpose") +
  ggtitle("Distribution by Loan Purpose")
```
```{r echo = FALSE}
ggplot(data = subset(pld2, LoanPurpose != "Not available"), 
       aes(x = LoanPurpose)) + 
  geom_bar(fill = "lightgreen") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_y_log10() + 
  ylab("Number of Loans") +
  xlab("Loan Purpose") +
  ggtitle("Distribution by Loan Purpose")
```

#### BORROWERS RATE:     

Next, I take a look at the base interest rate offered to borrowers. Right of the bat, I can see that there's a difference in the rates offered to borrowers. The minimum rate offered was 0% and the maximum rate offered was 49.75%.   I wonder what determines the rate of interest offered to a borrower?     

```{r echo=FALSE}
summary(pld2$BorrowerRate)
```
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = pld2, aes(x = BorrowerRate)) + 
  geom_histogram(fill = "lightskyblue", color = "black") + 
  geom_freqpoly() + 
  scale_x_continuous(breaks = seq(0, 0.4, 0.05)) + 
  ylab("Number of Loans") +
  xlab("Borrower Interest Rate") +
  ggtitle("Distribution of Loans by Borrower Rate")
```

I will look at the loan count over time.

To make it easier to work with dates, I'm going to store the Loan Origination Date in the yyy-mm-dd format.      
                
```{r echo=FALSE}
#converting variable to date format

pld2 <- pld2 %>% mutate(LoanOriginationDate = ymd(substr(LoanOriginationDate, 
                                                         1, 10)))
head(pld2$LoanOriginationDate)
```
           
    
At the first glance, I see that the plot of the number of loans made overtime has a gap - where no loans were made - from late 2008 to early 2009. This coincides with the unfurling of the Great Recession, when lender confidence brokedown and credit markets froze in the United States.         


```{r echo = FALSE}

#Grouping data by Origination Date

pld2_by_date <- pld2 %>% 
  group_by(LoanOriginationDate) %>% 
  summarize(count = n(), average_rate = mean(BorrowerRate))

ggplot(data = pld2_by_date, aes(x = LoanOriginationDate, y = count)) + 
  geom_point(alpha = 1/2, color = "midnightblue") + 
  ylab("Number of Loans") +
  xlab("Year") +
  ggtitle("Number of Loans over time")
```

# Univariate Analysis

### What is the structure of your dataset?

The dataset has 113937 observations, and 81 variables. 


### What is/are the main feature(s) of interest in your dataset?       

The primary feature of interest in my data set is the interest rate: Borrower Rate. I observed that borrowers were offered different interest rates. I am interested in understanding what goes into calculating the rate of interest.         

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?       

I know that the interest rate depends on how much of a risk the borrower is considered to be. A higher risk means a higher interest rate. So I will look at features that help me evaluate the risk profile.       

These include:      

Discrete: Income Range, Employment Status, Loan Purpose
Continuous: Monthly Income, Credit Score (Mean), Loan Amount      

### Did you create any new variables from existing variables in the dataset?     

Yes, I calculated the Mean Credit Score from the Upper and Lower ranges of borrowers credit score.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?       

I used a log10 tranformation on the y-axis of plots where the data was either long tailed or had high variance to view the distibution better.        
         
One plot that stood out was that of StatedMonthlyIncome. The distribution was not very telling because of the presence of outliers. I limited the axes to get a zoom in on the bulk of the data.      

------------

# Bivariate Plots Section




When I plot the borrowers rate over time, I get a very noisy plot. This is because we have data per day, and the daily average tends to move around a lot. To get a clearer picture, I plot the average rate per year, this smoothens the curve out. It looks like the rates were on  the rise till 2011, and then begin a gradual fall.     

```{r echo=FALSE}
ggplot(data = pld2_by_date, aes(x = LoanOriginationDate, y = average_rate)) + 
  geom_point(alpha = 0.5, color = "midnightblue") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) + 
  ylab("Borrower Rate") +
  xlab("Year") +
  ggtitle("Borrower Rate over Time")
```

---

From the above two graphics, I see that there exists a negative relationship between the Number of Loans taken out and the interest rate. This is an intuitive fact: As the cost of borrowing goes up, less number of people borrow.   
         
I would like to quantify this relationship in the Prosper Data Set.         

---
             
```{r echo=FALSE}

cor.test(pld2_by_date$average_rate, pld2_by_date$count)

```

---

The coefficient is neglgible.

----

I will continue with my exploration of the Borrower Rate.      
              
On plotting the daily average borrower rate,  I get a very noisy plot - which is typical interest rate behaviour. But, not much is discernible from this plot. Except, I see that the borrower rate was excessively volatile in the period from mid 2009 - 2011.                  
          
To get a sense of the long-run trend in the borrower rate, I plot the yearly average.        
              
```{r echo=FALSE}
t1 <- ggplot(data = pld2_by_date, 
             aes(x = LoanOriginationDate, y = average_rate)) + 
  geom_line(stat = "summary", fun.y = median, color = "firebrick") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  ylab("Borrower Rate") +
  xlab("Year") +
  ggtitle("Borrower Rate over Time")

#Plotting data per year

t2 <- ggplot(data = pld2_by_date, aes(x = year(LoanOriginationDate), 
                                      y = average_rate)) + 
  geom_line(stat = "summary", fun.y = median, color = "firebrick") + 
  scale_x_continuous(breaks = seq(2005, 2015, 1)) + 
  coord_cartesian(xlim = c(2006, 2014)) +
  ylab("Borrower Rate") +
  xlab("Year")

grid.arrange(t1, t2) 
```
         
Next I plotted the average borrower rates year-on-year and quarter-on-quarter. I see wave like functions emerging for both the plots.       
            
```{r echo=FALSE, Bivariate_Plots}
average.BorrowerRate <- pld2 %>% group_by(LoanOriginationQuarter) %>% 
  summarize(avg.rate = mean(BorrowerRate))

p1 <- qplot(data = average.BorrowerRate, x = LoanOriginationQuarter, 
            y = avg.rate, main = 'Year on Year') + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_x_discrete() +
  ylab("Borrower Rate") +
  xlab("Time (Quarter)") +
  ggtitle("Average Borrowe Rate over Time")

#Changing period data into Year-Quarter format.

o <- substr(average.BorrowerRate$LoanOriginationQuarter, 1, 2)
t <- substr(average.BorrowerRate$LoanOriginationQuarter, 4, 7)

average.BorrowerRate$quarter.Transformed <- paste(t, o)
a <- average.BorrowerRate %>% arrange(quarter.Transformed)

p2 <- ggplot(data = a, aes(x = quarter.Transformed, 
                           y = avg.rate, group = 1)) + 
  geom_line(color = "blue") + geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Borrower Rate") +
  xlab("Time (Quarter)") +
  ggtitle("Quarter on Quarter")

grid.arrange(p1, p2)
```



#### BORROWER RATE BY EMPLOYMENT STATUS

Looking at the below charts, I can see that people who are unemployed tend to get a higher rate of interest. This is consistent because interest rate is a compensation for risk, and unemployed people pose a higher risk for lenders.

```{r echo=FALSE}
ggplot(data = subset(pld2, !is.na(EmploymentStatus) & 
                       EmploymentStatus != "Not available"), 
       aes(x = EmploymentStatus, y = BorrowerRate)) + 
         geom_jitter(alpha = 1/20, color = "skyblue") + 
         scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  xlab("Employment Status") + ylab("Borrower Rate") +
  ggtitle("Average Borrower Rate by Employment Status")
```


```{r echo=FALSE}
ggplot(data = subset(pld2, !is.na(EmploymentStatus) & 
                       EmploymentStatus != "Not available"), 
       aes(x = EmploymentStatus, y = BorrowerRate)) + 
  geom_point(stat = "summary", fun.y = mean, size = 5, 
             color = "lightblue") +
  xlab("Employment Status") + ylab("Borrower Rate") +
  ggtitle("Average Borrower Rate by Employment Status")
```


```{r echo=FALSE}
ggplot(data = subset(pld2, !is.na(EmploymentStatus) & 
                       EmploymentStatus != "Not available"), 
       aes(x = EmploymentStatus, y = BorrowerRate)) + 
  geom_boxplot(color = "lightblue") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  xlab("Employment Status") + ylab("Borrower Rate") +
  ggtitle("Average Borrower Rate by Employment Status")
```
                            
#### HOMEOWNERSHIP AND BORROWER RATE:                           
                
I see that people who were not homeowners were stuck with higher borrower rates on average compared to homeowners. This is probably because homeowners have a stable security as collateral.       
           
```{r echo = FALSE}
ggplot(data = subset(pld2, !is.na(IsBorrowerHomeowner) & 
                       IsBorrowerHomeowner != "Not available"), 
       aes(x = IsBorrowerHomeowner, y = BorrowerRate)) + 
  geom_jitter(alpha = 1/35, color = "lightpink") + 
  scale_y_continuous(breaks = seq(0, 0.5, 0.05)) +
  ggtitle("Interest Rate by Home Ownership Status") +
  xlab("Home Ownership") + ylab("Borrower Rate") 
```

```{r echo = FALSE}
ggplot(data = subset(pld2, !is.na(IsBorrowerHomeowner) & 
                       IsBorrowerHomeowner != "Not available"), 
       aes(x = IsBorrowerHomeowner, y = BorrowerRate)) + 
  geom_point(stat = "summary", fun.y = mean, size = 5, 
             color = "lightpink") +
  ggtitle("Mean Interest Rate by Home Ownership Status") +
  xlab("Employment Status") + ylab("Borrower Rate") 

```

```{r echo = FALSE}
ggplot(data = subset(pld2, !is.na(IsBorrowerHomeowner) & 
                       IsBorrowerHomeowner != "Not available"), 
       aes(x = IsBorrowerHomeowner, y = BorrowerRate)) + 
  geom_boxplot() + 
  ggtitle("Interest Rate by Home Ownership Status") +
  xlab("Home Ownsership") + ylab("Borrower Rate") 
```


#### LOAN AMOUNT, CREDIT SCORES, INCOME AND BORROWER RATE:        
      
Below are the continuous variables under consideration. I plotted them in a scatterplot matrix.    
       
I see that Borrower rate has negligible negative correlation with the Loan Amount, Income and Credit Score.    

Maybe the Borrower Rate is determined by a combination of these factors, and taken alone neither of these variables can explain away Borrower Rates.      
         
```{r echo=FALSE, fig.height=10, fig.width=10}
for_cor_matrix <- pld2[,c("LoanOriginalAmount", "CreditScoreMean", 
                          "StatedMonthlyIncome", "BorrowerRate")]
ggpairs(subset(for_cor_matrix, StatedMonthlyIncome <= 50000 & 
                 CreditScoreMean > 400))
```
             
#### WHAT DOES PROSPER ASK FOR A CUSTOM RATE?         

I went to Prosper's website to  see how exactly the borrower rate is determined. Prosper prompts the user to enter information relating to four variables in our data set:     
      
1. The Loan Amount required
2. Borrower's Income
3. Employment Status
4. Purpose of the Loan        
         
I have already explored these variables above, but I think it would be telling to plot them in one matrix.        
                 

```{r echo = FALSE, fig.height=10, fig.width=10}
Prosper_requirement <- pld2[, c('LoanOriginalAmount', 'StatedMonthlyIncome', 
                                'EmploymentStatus', 'LoanPurpose', 
                                'BorrowerRate')]

# Changed cardinality threshold to accomodate LoanPurpose

ggpairs(subset(Prosper_requirement, (LoanPurpose != c("Not available", 
                                                      "RV", "Green loans", 
                                                      "Engagement ring", 
                                                      "Cosmetic procedure", 
                                                      "Boat", 
                                                      "Baby and adoption")) & 
                 (StatedMonthlyIncome <= 50000)), cardinality_threshold = 21)
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?            

I see that the Credit Score, Income and Loan Amount have negligible negative correlation with the Borrowers Rate. I think a combination of these factors will give us a more discernable relationship.       
        
I also observed that the Borrowers Rate - on average - came down gradually over the years. Higher risk borrowers like the unemployed or those who are not homeowners are stuck with a higher rate of interest.
                    

--------------

# Multivariate Plots Section     

I trained two linear models:     

One with Loan Amount, Income, Employment Status and Purpose of the loan - this model explained only 10% of the variance.   
I updated this model with the Credit Score and was able to explain 30% of the variance.      
```{r echo = FALSE}
model_1 <- lm(BorrowerRate ~ LoanOriginalAmount + StatedMonthlyIncome + 
                EmploymentStatus + LoanPurpose, data = pld2)

model_2 <- update(model_1, . ~ . + CreditScoreMean)

mtable(model_1, model_2)
```
      
Given that the models above were only able to explain 30% of the variation in the borrower's rate, I decided to look for additional factors that might affect the determination of borrowers rate.        

This time I decide to rely on Prospers measure of credit risk: Prosper Score. Prosper Score is calculated by taking into account repayment history of previously taken loans.   

```{r echo=FALSE, Multivariate_Plots}

pld2$ProsperScore <- pld$ProsperScore

summary(pld2$ProsperScore)
```
             
Since Prosper Scores were awarded only after May-2009, I will only be looking at data from this time frame.

```{r}
pld2$LoanOriginationDate <- ymd(pld2$LoanOriginationDate)

to_plot_1 <- subset(pld2, year(LoanOriginationDate) > 2009 & 
                    !is.na(ProsperScore))

color_count <- length(unique(to_plot_1$ProsperScore))
get_palette <- colorRampPalette(brewer.pal(9, "Set1"))

ggplot(data = to_plot_1, aes(x = LoanOriginationDate, y = BorrowerRate)) + 
  geom_point(aes(color = factor(ProsperScore))) +
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) + 
  ggtitle("Borrower Rate over Time and Prosper Score") +
  ylab("Borrower Rate") +
  xlab("Year") +
  labs(color = "Prosper Score")
```



```{r echo = FALSE, warning=FALSE, message=FALSE}
#updating grouped data frame

pld2_by_date <- pld2 %>% 
  group_by(LoanOriginationDate) %>% 
  summarize(count = n(), average_rate = mean(BorrowerRate), 
            average_prosper_score = mean(ProsperScore), 
            credit_rating = mean(CreditScoreMean))


#subsetting to remove NAs
#Prosper Scores were given out only after 2009 - subsetting data to include this

to_plot <- subset(pld2_by_date, year(LoanOriginationDate) > 2009 & 
                    !is.na(average_prosper_score))

ggplot(data = to_plot, aes(x = LoanOriginationDate, y = average_rate)) + 
  geom_point(aes(color = average_prosper_score)) + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) + 
  geom_smooth() + 
  ggtitle("Mean Borrower Rate over Time and Prosper Score") +
  ylab("Borrower Rate") + 
  xlab("Year")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = to_plot, aes(x = average_prosper_score, y = average_rate)) + 
  geom_point(aes(color = year(LoanOriginationDate))) + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) + 
  xlab("Prosper Score") +
  ylab("Borrower Rate") + 
  ggtitle("Borrower Rate vs Prosper Score over Time") +
  labs(color="Year")
```
           
From a couple of plots I can see that the Prosper Score awarded has gone down - on average - over time. I also observed clustering of observations.   
        
         
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = to_plot, aes(x = average_prosper_score, y = average_rate)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  ggtitle("Prosper Score vs. Borrower Rate") +
  xlab("Prosper Score") +
  ylab("Borrower Rate")
  
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = to_plot, aes(x = credit_rating, y = average_rate)) + 
  geom_point() + geom_smooth(method = "lm") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  ggtitle("Credit Score vs. Borrower Rate") +
  xlab("Credit Score") +
  ylab("Borrower Rate")
```
             
The two plots above seem to have a better structure than the plots in the previous section, and I think it will be worthwhile to train linear models employing them.

```{r echo=FALSE}
model_3 <- lm(BorrowerRate ~ ProsperScore, data = pld2)
model_4 <- update(model_3, . ~ . + CreditScoreMean)

mtable(model_3, model_4)
```
         
I can see that I get better models this time around.       
            
When I looked at Borrower Rate vs Prosper Score over time, I found clustering. This indicates that Time could be an important factor explaining Borrower Rates.        
        
```{r echo=FALSE}
#making a new variable for year
pld2$Loan_origination_year <- year(pld2$LoanOriginationDate)

#making a continuous quarter variable
pld2$Loan_origination_quarter <- quarter(pld2$LoanOriginationDate, 
                                         with_year = TRUE)

model_5 <- update(model_4, . ~ . + Loan_origination_year)
model_6 <- update(model_4, . ~ . + Loan_origination_quarter)

mtable(model_5, model_6)
```

We get a slighty better model that explains about 60% of the variance in the Borrower Rate.      

We can fine-tune this model by looking at time in fiscal quarters. This results in a yet better model that explains about 70% of the variance in the Borrower Rate        

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest? Were there any interesting or surprising interactions between features?

I can see in the plot of Prosper Score vs Borrower Rate over Time that there are clusters of loans:      
The loans awarded 2013 and 2014 are concentrated and fall between 4 and 6 clocking borrower rates approximately between 15% and 22%;          
The loans awarded in 2012 are also concentrated and have scores between 5 and 7, with borrower rates between 17% and 25%;         
The scores and borrower rates for loans before 2012 are more spread out.           

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

Yes, I created a few models, best of which comes with an R-squared of 0.7. This explains 70% of the variation in the borrower rate. Which is a good metric to understand the anatomy of how Prosper borrower rates are determined.     

However, the data set was riddled with missing values, and the model was trained on a subset of the original data set. This could take away accuracy.      
       
All in all, the model is good to understand what determined borrowe rates - Prosper Score (Calculated based on past repayment record), Credit Score and the time of the year the loan was taken.      
      
A major limiting factor for this model is that it is not a very robust predictive model.      
               
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message = FALSE, warning = FALSE, Plot_One}
t1 <- ggplot(data = pld2_by_date, aes(x = LoanOriginationDate, 
                                      y = average_rate)) + 
  geom_line(stat = "summary", fun.y = median, color = "firebrick") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  ylab("Borrower Rate") + 
  xlab("Year") +
  ggtitle("Borrower Rate over Time")

#Plotting data per year

t2 <- ggplot(data = pld2_by_date, aes(x = year(LoanOriginationDate), 
                                      y = average_rate)) + 
  geom_line(stat = "summary", fun.y = median, color = "firebrick") + 
  scale_x_continuous(breaks = seq(2005, 2015, 1)) + 
  coord_cartesian(xlim = c(2006, 2014)) +
  ylab("Borrower Rate") +
  xlab("Year")

grid.arrange(t1, t2)
```

### Description One        

I made two plots: one, borrower rate over time, and two, the average borrower rates. The Borrower Rate is highly volatile, but we can see a smooth trend - the rate, on average, gradually climbed from 2007 - 2011, and began a noticeable decent from 2011 - 2014.


### Plot Two
```{r echo=FALSE, Plot_Two}
p3 <- ggplot(data = to_plot, aes(x = average_prosper_score, y = average_rate)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  ggtitle("Prosper Score vs. Borrower Rate") +
  xlab("Prosper Score") +
  ylab("Borrower Rate")

p4 <- ggplot(data = to_plot, aes(x = credit_rating, y = average_rate)) + 
  geom_point() + geom_smooth(method = "lm") + 
  scale_y_continuous(breaks = seq(0, 0.35, 0.05)) +
  ggtitle("Credit Score vs. Borrower Rate") +
  xlab("Credit Score") +
  ylab("Borrower Rate")

grid.arrange(p3, p4, nrow = 1)
```

### Description Two        
    
Of the variables explored, Prosper Score and the Credit Score do the best job of explaining the Borrower Rate. Prosper Score is a credit score assigned by Prosper to assess credit risk, and the credit score is also a risk metric. Put together, they were able to explain about 50% of the variation in the interest rates - with a R-squared of 0.5.

Since interest rates are a compensation for risk, these metrics do a neat job of explaining some variation.


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = to_plot, aes(x = average_prosper_score, y = average_rate, color = year(LoanOriginationDate))) + 
  geom_point() + 
  xlab("Prosper Score") +
  ylab("Borrower Rate") + 
  ggtitle("Borrower Rate vs Prosper Score over Time") +
  labs(color = "Year")

```

### Description Three     

Clustering by time reveals that time could help explain variation in the borrower rate.

I plotted Borrower Rate vs Prosper Score. I adopted an unconventional method of color coding data points by year to look for clustering. 

------

# Reflection       
       
The Prosper Loan Dataset consists of information relating to 113937 crowd-sourced loans spread over 81 variables from 2005 to 2014. The first thing I encountered whilst exploring the dataset was empty cells - which I had to replace with NAs. I then dropped variables with high number of NAs from the analysis.    
    
I went through the variable descriptions, and started plotting to get a sense of the borrower's profile and characteristics of the loans. The plot for borrowers rate caught my eye. I saw that each borrower got a custom rate, and proceeded to understand what determines the rate offered.       
          
On visiting the Prosper [website](https://www.prosper.com/borrower/#/prospect/registration) I found that Prosper promts you to enter information regarding the required Loan Amount, Employment Status, Purpose of the Loan and Income. I then went on to explore these variables and a few others.       
          
While I got an approximate of what could be driving the borrower rate in some instances (like Unemployed and Non-homeowners get offered higher rates on average), I was mostly stuck with weak relationships between the borrower rate and other variables. I hypothesized that the weak relationship was on account that any standalone variable could not explain the Borrower Rate, but maybe a combination of variables could - however, this looked unlikely given the bivariate plots.    
         
As expected, the linear model came with an R-squared value of 0.3. At this point, I decided to take a look at other variables that might explain the borrower rate better. I found these variables to be the Prosper Score, Credit Score and the time. The final model could explain 70% of the variance. However it's limitations are that it was based on a dataset riddled with missing values, leading to loss in accuracy. The model is not a robust predictive model.

I intend to take this analysis further by implementing Machine Learning algorithms - including methods like K-nearest neighbors, Support Vector Machines (Using RBF Kernels) and Random Forests - to explore the same variables, to better discern any relationships.
           

         
